Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2016-01-21 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
stronger conclusions, better selection of GW mesh plots

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent atmospheric dynamics over steep orography.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  In contrast, here a single C-grid finite volume model is used to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test starting from stably stratified initial conditions, and orographically induced gravity waves described by nonhydrostatic dynamics.  In addition, two new advection tests are formulated having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.  These new tests present a challenge to the advection scheme on cut cell grids.
The results of the advection tests demonstrate that accuracy depends primarily upon alignment of the flow with the grid rather than the grid orthogonality.  As expected, the cut cell grid maintains greater accuracy for horizontal advection and in the test of stationary, stratified flow.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature, although terrain following velocity fields lead to errors on cut cell grids.
Due to semi-implicit timestepping and an upwind-biased, explicit advection scheme, there were no timestep restrictions associated with small cut cells.}

@@

The domain width is \SI{301}{\kilo\meter}, taken as the horizontal distance between the inlet and outlet boundaries.  The domain is \SI{25}{\kilo\meter} high, discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  Note that \citet{schaer2002} measured the domain width as \SI{300}{\kilo\meter} between the outermost cell centres where tracer values are specified.  The two domain sizes are equivalent irrespective of the measure.

@@

The test was executed on the BTF, SLEVE and cut cell grids, and on a regular grid with flat terrain using a centred linear scheme and the upwind-biased cubic scheme.   Results were also obtained on BTF and SLEVE grids with the fourth order scheme from \citet{schaer2002} using the modified version of their code.

@@

The results of the cubic upwind-biased scheme on TF and regular grids are comparable with those for the fourth-order centred scheme from \citet{schaer2002}.  Error is largest on the BTF grid with \(\ell_2 = \num{0.112}\) but is significantly reduced on the SLEVE grid with \(\ell_2 = \num{0.0146}\).
Advection is most accurate on the cut cell and regular grids, with \(\ell_2\) approximately half of that on the SLEVE grid.
Tracer minima and maxima for the centred linear and fourth order schemes are lower than those presented by \citet{schaer2002} because no interpolation is used to calculate the streamfunction.

The results of the horizontal advection test show that numerical errors are due either to misalignment of the flow with the grid, or to grid distortions.  Using the upwind-biased cubic scheme, distortions in the grid do not significantly distort the tracer.

@@

The simulation is integrated forward by 5 hours and the timestep, $\Delta t = 8 \Delta z / \SI{300}{\second}$, is chosen so that it scales linearly with spatial resolution and, following the original test specified by \citet{schaer2002}, $\Delta t = \SI{8}{\second}$ when $\Delta z = \SI{300}{\meter}$.
Test results are compared between the BTF and cut cell grids at several resolutions.
The spatial and temporal resolutions tested are shown in table~\ref{tab:gw-resolutions}.  The lowest resolution is the same as that used by \citet{schaer2002}, and higher resolutions preserve the same aspect ratio.
The vertical resolution is chosen to test various configurations of cut cell grid.  At $\Delta z = \SI{300}{\meter}$, the mountain lies entirely within the lowest layer of cells, while at $\Delta z = \SI{250}{\meter}$ and $\Delta z = \SI{125}{\meter}$ the mountain peak is aligned with the interface between layers.  With increasing resolutions up to $\Delta z = \SI{50}{\meter}$, the orography intersects more layers and becomes better resolved.  Three of the cut cell grids are shown in figure~\ref{fig:gw-meshes} at $\Delta z = \SI{300}{\meter}$, $\SI{250}{\meter}$ and $\SI{200}{\meter}$.  Small cells are visible on the $\SI{200}{\meter}$ grid but, on the $\SI{250}{\meter}$ grid, the grid generator has merged the small cells with those in the layer above.

@@

\subsection{Terrain following advection of thermal profile}
The potential temperature anomalies in the gravity waves test varied between terrain following and cut cell grids of the same resolution.  This variation may be due to differences in the wind fields between grids, or errors in the advection of potential temperature, amongst other possible causes.  To help establish the primary source of error, a new advection test is formulated in which the initial potential temperature field from the gravity waves test is used.  To eliminate any differences in wind fields, the field is advected in a fixed, terrain-following velocity field that mimics the flow in the gravity waves test.

@@

For comparison with the potential temperature anomalies in the gravity waves test, vertical profiles of potential temperature error are taken at $x = \SI{50}{\kilo\meter}$.  As seen in figure~\ref{fig:gw-theta}c, errors are negligible on the BTF grids, but figure~\ref{fig:gw-theta}d reveals significant errors in the lowest layers of the cut cell grids that were advected from the mountain peaks.  \TODO{the following is no longer true.  in fact, so far, the reverse appears to be true: errors are larger in the GW tests compared to the thermal advection tests ... ``The magnitude of error is about ten times larger than the potential temperature anomalies found on the cut cell grids in the gravity waves test seen in figure~\ref{fig:gw-theta}b.  We offer two explanations for this difference: first, there are no restoring forces in the linear advection model and, second, the wind fields in the two tests are different.''}

@@

Across all tests, a high degree of accuracy was achieved for all grids.  Even on the highly-distorted BTF grid, which has previously been found to give poor results \citep{schaer2002,klemp2011,good2014}, errors were often small in the tests presented here.  In the first two tests, tracers were advected by horizontal and terrain following velocity fields.  We found that the accuracy of the upwind-biased cubic advection scheme depended upon alignment of the flow with the grid rather than on grid distortions.
Spurious vertical velocities in the resting atmosphere test were similar on terrain following and cut cell grids.
In the gravity waves test, vertical velocities were in good agreement with the reference solution from \citet{melvin2010} across all grids.

@@

The cubic upwind-biased advection scheme takes an approach for treating small cut cells that differs from other existing approaches by adjusting weightings to ensure that advection remains upwind-biased near small cells.  The implementation of this technique in OpenFOAM is available at \url{https://github.com/hertzsprung/AtmosFOAM/tree/shaw-weller-2015-mwr} and will be described in greater detail a future publication.  Combined with semi-implicit timestepping and a grid generator that preserves cell length in the direction of the flow, small cells did not impose additional timestep constraints.

@@

\begin{figure*}
	\centering
	\includegraphics{../fig-grid-generation/fig-grid-generation.pdf}
	%
	\caption{Illustration of a slanted cell grid (a) before, and (b) after construction.
	The terrain surface, denoted by a heavy dotted line, intersects a uniform rectangular grid comprising six cells, $c_1, \ldots, c_6$.  The cell vertices, marked by open circles, are moved to the points at which the terrain intersects vertical cell edges, marked by filled circles.  Cells that have no volume are removed.  Where a cell has two vertices occupying the same point, the zero-length edge that joins those vertices is removed.  In this illustration, cells $c_5$ and $c_6$ are removed because they have no volume, and the zero-length edge at point $q$ is removed to create a triangular cell, $c_4$.  Point $p$ is moved down because it is within $2 \Delta z/5$ of the surface, avoiding the creation of a thin cell.}
	\label{fig:grid-generation}
\end{figure*}

@@

\begin{figure*}
	\centering
	\includegraphics{../fig-gravityWaves-meshes/fig-gravityWaves-meshes.pdf}
	%
	\caption{Cut cell grids used for the gravity waves and thermal advection tests at (a) $\Delta z = \SI{300}{\meter}$, (b) $\Delta z = \SI{250}{\meter}$, and (c) $\Delta z = \SI{200}{\meter}$.  The mountain peak $h_0 = \SI{250}{\meter}$.  At $\Delta z = \SI{250}{\meter}$, the grid creation process has merged small cells with the cells in the layer above but, at $\Delta z = \SI{200}{\meter}$, small cells have been retained.  The full two dimensional grids are \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high.  Axes are in units of \si{\meter}.}
	\label{fig:gw-meshes}
\end{figure*}

>>>>

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent atmospheric dynamics over steep orography.
Because the type of grid is usually chosen during model implementation, it becomes necessary to use different models to compare the accuracy of different grids.  In contrast, here a C-grid finite volume model enables a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test starting from resting initial conditions, and orographically induced gravity waves described by nonhydrostatic dynamics.
In addition, three new tests are formulated: a more challenging resting atmosphere case, and two new advection tests having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.  These new tests present a challenge on cut cell grids.  
The results of the advection tests demonstrate that accuracy depends primarily upon alignment of the flow with the grid rather than grid orthogonality.  A resting atmosphere is well-maintained on all grids.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature, although terrain following velocity fields lead to errors on cut cell grids.
Due to semi-implicit timestepping and an upwind-biased, explicit advection scheme, there are no timestep restrictions associated with small cut cells.
We do not find the significant advantages of cut cells or smoothed coordinates that other authors find.}

@@

The domain width is \SI{301}{\kilo\meter}, taken as the horizontal distance between the inlet and outlet boundaries.  The domain is \SI{25}{\kilo\meter} high, discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  Note that \citet{schaer2002} measured the domain width as \SI{300}{\kilo\meter} between the outermost cell centres where tracer values are specified.
Both formulations create a cell centre (or mass point) rather than a cell face (or horizontal velocity point) over the top of the highest peak which is crucial for the accuracy of the centred advection schemes.

@@

The test was executed on the BTF, SLEVE and cut cell grids using a centred linear scheme and the upwind-biased cubic scheme.   Results were also obtained on BTF and SLEVE grids with the fourth order scheme from \citet{schaer2002} using the modified version of their code.

@@

The results of the cubic upwind-biased scheme on TF and regular grids are comparable with those for the fourth-order centred scheme from \citet{schaer2002}.  Error is largest on the BTF grid with \(\ell_2 = \num{0.112}\) but is significantly reduced on the SLEVE grid with \(\ell_2 = \num{0.0146}\).
Advection is most accurate on the cut cell grid, with \(\ell_2\) approximately half of that on the SLEVE grid.
Tracer minima and maxima for the centred linear and fourth order schemes are lower than those presented by \citet{schaer2002} because no interpolation is used to calculate the streamfunction.

The results of the horizontal advection test show that numerical errors are due either to misalignment of the flow with the grid, or to grid distortions.
In the following section, we propose a new test in order to identify the cause of the errors.

@@

The simulation is integrated forward by 5 hours and the timestep, $\Delta t = 8 \Delta z / \SI{300}{\second}$, is chosen so that it scales linearly with spatial resolution and, following the original test specified by \citet{schaer2002}, $\Delta t = \SI{8}{\second}$ when $\Delta z = \SI{300}{\meter}$.
Test results are compared between the BTF and cut cell grids at several resolutions.
The spatial and temporal resolutions tested are shown in table~\ref{tab:gw-resolutions}.  The lowest resolution is the same as that used by \citet{schaer2002}, and higher resolutions preserve the same aspect ratio.
The vertical resolution is chosen to test various configurations of cut cell grid.  At $\Delta z = \SI{300}{\meter}$, the mountain lies entirely within the lowest layer of cells, while at $\Delta z = \SI{250}{\meter}$ and $\Delta z = \SI{125}{\meter}$ the mountain peak is aligned with the interface between layers.  With increasing resolutions up to $\Delta z = \SI{50}{\meter}$, the orography intersects more layers and becomes better resolved.  Three of the cut cell grids are shown in figure~\ref{fig:gw-meshes} at $\Delta z = \SI{300}{\meter}$, $\SI{200}{\meter}$ and $\SI{150}{\meter}$.  Small cells are visible on the $\SI{150}{\meter}$ grid but, on the $\SI{200}{\meter}$ grid, the small cells are merged with those in the layer above.

@@

\subsection{Terrain following advection of thermal profile}
The potential temperature anomalies in the gravity waves test do not converge with resolution when using the cut cell grids.  This may be due to differences in the wind fields between grids, or errors in the advection of potential temperature, amongst other possible causes.  To help establish the primary source of error, a new advection test is formulated in which the initial potential temperature field from the gravity waves test is used.  To eliminate any differences in wind fields, the field is advected in a fixed, terrain-following velocity field that mimics the flow in the gravity waves test.

@@

For comparison with the potential temperature anomalies in the gravity waves test, vertical profiles of potential temperature error are taken at $x = \SI{50}{\kilo\meter}$.  As seen in figure~\ref{fig:gw-theta}c, errors are negligible on the BTF grids, but figure~\ref{fig:gw-theta}d reveals significant errors in the lowest layers of the cut cell grids that were advected from the mountain peaks.

@@

Across all tests, a high degree of accuracy was achieved for all grids.  Even on the highly-distorted BTF grid, which has previously been found to give poor results \citep{schaer2002,klemp2011,good2014}, errors were often small in the tests presented here.  In the first two tests, tracers were advected by horizontal and terrain following velocity fields.  We found that the accuracy of the upwind-biased cubic advection scheme depended upon alignment of the flow with the grid rather than on grid distortions.
Spurious vertical velocities in the resting atmosphere tests were similar on terrain following and cut cell grids.
In the gravity waves test, vertical velocities were in good agreement with the reference solution from \citet{melvin2010} across all grids.

@@

The cubic upwind-biased advection scheme takes an approach for treating small cut cells that differs from other existing approaches by adjusting weightings to ensure that advection remains upwind-biased near small cells.  The implementation of this technique in OpenFOAM is available at \url{https://github.com/hertzsprung/AtmosFOAM/tree/shaw-weller-2015-mwr} and will be described in greater detail a future publication.  Combined with semi-implicit timestepping and a new cut cell generation technique that preserves cell length in the direction of the flow, small cells did not impose additional timestep constraints.
By using a suitable multidimensional advection scheme and a curl-free pressure gradient formulation, we did not find significant advantages of cut cells or smoothed coordinate systems unlike \citet{good2014,klemp2011,schaer2002}.
In contrast, errors that do not reduce with resolution are on cut cell grids.  No significant problems were found when using basic terrain following grids.

@@

\begin{figure*}
	\centering
	\includegraphics{../fig-grid-generation/fig-grid-generation.pdf}
	%
	\caption{Illustration of a cut cell grid (a) before, and (b) after construction.
	The terrain surface, denoted by a heavy dotted line, intersects a uniform rectangular grid comprising six cells, $c_1, \ldots, c_6$.  The cell vertices, marked by open circles, are moved to the points at which the terrain intersects vertical cell edges, marked by filled circles.  Cells that have no volume are removed.  Where a cell has two vertices occupying the same point, the zero-length edge that joins those vertices is removed.  In this illustration, cells $c_5$ and $c_6$ are removed because they have no volume, and the zero-length edge at point $q$ is removed to create a triangular cell, $c_4$.  Point $p$ is moved down because it is within $2 \Delta z/5$ of the surface, avoiding the creation of a thin cell.}
	\label{fig:grid-generation}
\end{figure*}

@@

\begin{figure*}
	\centering
	\includegraphics{../fig-gravityWaves-meshes/fig-gravityWaves-meshes.pdf}
	%
	\caption{Cut cell grids used for the gravity waves and thermal advection tests at (a) $\Delta z = \SI{300}{\meter}$, (b) $\Delta z = \SI{200}{\meter}$, and (c) $\Delta z = \SI{150}{\meter}$.  The mountain peak $h_0 = \SI{250}{\meter}$.  At $\Delta z = \SI{300}{\meter}$ and $\Delta z = \SI{200}{\meter}$, the grid creation process has merged small cells with the cells in the layer above but, at $\Delta z = \SI{150}{\meter}$, small cells have been retained.  The full two dimensional grids are \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high.  Axes are in units of \si{\meter}.}
	\label{fig:gw-meshes}
\end{figure*}
